WSClean: an implementation of a fast, generic wide-field imager

### 2014-02-10 articles/2014/Article-WSClean/wsclean.tex
Updating snapshot example image with vectorized one

There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking\footnote{TODO when/where was this first mentioned in literature?} \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}. The w-snapshots algorithm is a hybrid between snapshot imaging and one of the other w-term correcting techniques, most efficiently w-projection or w-stacking.

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). At the MWA, some of the data is reduced with the w-projection algorithm implementation of the Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}). This is an attractive option because of its many features, high quality and free availability. However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm. Another option for reducing MWA data is the Real-Time System (RTS), which has been designed as an efficient calibrating and imaging pipeline specifically for MWA data \citep{rts-mwa}. It can use GPUs for improving efficiency, and uses a w-projection kernel during imaging to correct for the w-term. The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to Zenith. The RTS can not perform deconvolution as of yet.

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for widefield telescopes. One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
One such DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure (TODO cite). This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.

Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge. In \citet{widefield-imaging-ska-cornwell}, it is shown that in theory the w-snapshots algorithm is the most efficient approach for the SKA.

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation. To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms, optionally combined with w-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''. This work improves the field of wide-field imaging in the following ways:
\begin{itemize}
 \item The performance of w-stacking is compared to w-projection.
 \item Parameters that affect the w-snapshots algorithm are analysed.
 \item Details are described of how to efficiently implement the w-stacking and w-snapshots algorithms.
 \item A new implementation is released, which we believe to be the fastest generic widefield imager available.
\end{itemize}

The w-stacking algorithm is described in Sect.~\ref{sec:wstacking}. Details of implementing the w-stacking and w-snapshots algorithms are described in Sect.~\ref{sec:implementation}. The performance and accuracy will be analysed in Sect.~\ref{sec:analysis}.

\section{The w-stacking technique} \label{sec:wstacking}
In this section, we will describe the w-stacking algorithm from a mathematical point of view.

An interferometer samples the complex visibility function
\begin{align}\notag
V(u,v,w) = & \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} \cdot \\ \label{eq:visibility-function}
& e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
\end{align}
where $u,v,w$ is a baseline coordinate in the coordinate system of the array, $A$ is the primary-beam function, $I$ is the sky function and $l,m$ are cosine sky coordinates. We will use $I'(l,m)$ to denote the sky function before primary-beam correction, $I'(l,m)=A(l,m)I(l,m)$. We will not discuss calibration, but assume $V$ has been calibrated before imaging. In the case of a polarized measurement, the symbols $V$, $A$ and $I$ are $2\times 2$ Jones matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem. Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.

@@

\subsection{Discretization of w} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. WSClean was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal TODO in contrast to some article? -- could be explained by having more samples at smaller $w$s. From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlayers} \gg 2\pi \left(w_e - w_0\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_e - w_0$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_e - w_0$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. [TODO table of arrays with $w$ values at zenith/more angles/more freq?] For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that, in the case of the MWA, tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\section{The WSClean imager implementation} \label{sec:implementation}
With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm. The imager is not specialized for the MWA, and has been succesfully used for other telescopes (TODO menton more).

@@

\subsection{W-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. TODO time difference and more discussion, replace with deeper images.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

@@

Table~\ref{tbl:accuracy-measurements} lists the measured standard errors of the source brightnesses as detected by Aegean and the RMS in the residual image. In all tests, WSClean performs slightly more accurate. At zenith, CASA produces a residual RMS twice as high as WSClean. This is caused by the fact that we keep the number of w-projection planes in w-projection equal to the number of w-layers in w-stacking, resulting in only 12 $w$-planes at zenith. This evidently has a stronger effect on the w-projection algorithm. However, the computational performance of the w-projection algorithm is hardly affected by the number of w-projection planes, and in practical situations one would always use more w-projection planes\footnote{In the zenith case CASA actually issues a warning to use more $w$-projection planes.}. When 128 w-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers. We do not know why this parameter needs to be higher in CASA to reach the same RMS. In theory it should be enough, since Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree. Also unexpected is that the source flux density becomes worse by increasing the number of planes. For WSClean both values stay approximately the same when the number of $w$-layers is increased.

@@

@@

>>>>

There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}. The w-snapshots algorithm is a hybrid between snapshot imaging and one of the other w-term correcting techniques, most efficiently w-projection or w-stacking.

New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}). The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) has an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution. However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm. Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.

Another option for reducing MWA data is the Real-Time System (RTS), which has been designed as an efficient calibrating and imaging pipeline specifically for MWA data \citep{rts-mwa}. It can use GPUs for improving efficiency, and uses a w-projection kernel during imaging to correct for the w-term. The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to Zenith. The RTS can not perform deconvolution as of yet.

To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs). This is especially true for widefield telescopes. One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
One such DDE is the effect of the ionosphere. For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}. This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data. The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.

Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge. \citet{widefield-imaging-ska-cornwell} shows that in theory the w-snapshots algorithm is the most efficient approach for the SKA.

In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation, and is able to succesfully image any MWA observation. To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms, optionally combined with w-snapshot imaging. We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''. Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{TODO add URL}.

This paper is structured as follows: The $w$-stacking algorithm is described in Sect.~\ref{sec:wstacking}. Details of implementing the $w$-stacking and $w$-snapshots algorithms are described in Sect.~\ref{sec:implementation}. The performance and accuracy will be analysed in Sect.~\ref{sec:analysis}.

\section{The w-stacking technique} \label{sec:wstacking}
In this section, we will describe the $w$-stacking algorithm from a mathematical point of view.

An interferometer samples the complex visibility function
\begin{align}\notag
V(u,v,w) = & \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} \cdot \\ \label{eq:visibility-function}
& e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
\end{align}
where $u,v,w$ is a baseline coordinate in the coordinate system of the array, $A$ is the primary-beam function, $I$ is the sky function and $l,m$ are cosine sky coordinates. We will use $I'(l,m)$ to denote the sky function before primary-beam correction, $I'(l,m)=A(l,m)I(l,m)$. We will not discuss calibration, but assume $V$ has been calibrated before imaging. In the case of a polarized measurement, the symbols $V$, $A$ and $I$ are $2\times 2$ matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem. Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.

@@

\subsection{Discretization of w} \label{sec:gridding-w}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/aliasing-example-annotated}
\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field. WSClean was set to use 12 $w$-layers. The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers. Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy. Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources. }
\label{fig:aliasing-example}
\end{center}
\end{figure}
While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed. For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian. This results in the constraint
\begin{equation} \label{eq:minimum-w-distance}
\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
\end{equation}
This suggest that a uniform discretization in $w$ is optimal (TODO this is in contrast to what Cornwell 2008 says -- mailed Cornwell to ask why). From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
\begin{equation} \label{eq:nwlayers-bound}
 N_\textrm{wlayers} \gg 2\pi \left(w_e - w_0\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
\end{equation}
Actual values for the right-hand side can differ a lot. The value of $w_e - w_0$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image. For the MWA, a typical value of $w_e - w_0$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$. For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$. This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations. The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.

@@

\section{The WSClean imager implementation} \label{sec:implementation}
With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm. As will be shown in Sect.~\ref{sec:application-to-nonmwa}, the imager is not specialized for the MWA, and has been succesfully used for imaging data from other telescopes.

@@

\subsection{W-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. TODO time difference and more discussion.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

@@

Table~\ref{tbl:accuracy-measurements} lists the measured RMS in the residual image and the standard errors of the source brightnesses as detected by Aegean. In all tests, WSClean performs slightly more accurate. At zenith, CASA produces a residual RMS twice as high as WSClean. This is caused by the fact that we keep the number of w-projection planes in w-projection equal to the number of w-layers in w-stacking, resulting in only 12 $w$-planes at zenith. This evidently has a stronger effect on the w-projection algorithm. However, the computational performance of the w-projection algorithm is hardly affected by the number of w-projection planes, and in practical situations one would always use more w-projection planes\footnote{In the zenith case CASA actually issues a warning to use more $w$-projection planes.}. When 128 w-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers. We do not know why this parameter needs to be higher in CASA to reach the same RMS. In theory it should be enough, since Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree. Also unexpected is that the source flux density becomes worse by increasing the number of planes. For WSClean both values stay approximately the same when the number of $w$-layers is increased.

@@

\section{Application of WSClean to other telescopes} \label{sec:application-to-nonmwa}
TODO

@@

TODO discuss optimalisations with GPUs
